---
title: "Models for Longfin Inshore Squid"
output: github_document
---

```{r knit, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
###### Zoe Ahmiegbe

### Steps
```{r setup, include=FALSE}
source("/home/zeahmi28/ColbyForecasting/setup.R")
```

#### Loading Data
To build a model, I need two things: the configuration file and covariate data (variables).
```{r load_data, include=FALSE}
cfg = read_configuration(scientificname = "Doryteuthis pealeii", version = "v1")
model_input = read_model_input(scientificname = "Doryteuthis pealeii", 
                               version = "v1",
                               log_me = c("depth", "Xbtm")) |>
  dplyr::mutate(month = month_as_number(.data$month)) |>
  select(all_of(c("class", cfg$keep)))
```

#### Splitting the Data
Why do I split the data? So the model doesn't just memorize what we give it. It's important that it's tested on unseen data. I'll use 80% of the data to train the model and will reserve 20% of the data for later testing.

```{r initial_split, include=FALSE}
model_input_split = spatial_initial_split(model_input, 
                        prop = 1 / 5,     # 20% for testing
                        strategy = spatial_block_cv) # see ?spatial_block_cv
model_input_split
```

This plot shows the spatial distribution of the training and testing data.
```{r initial_split_plot, echo=FALSE}
autoplot(model_input_split)
```

I then divide the training data into spatial cross-validation folds to help the model learn patterns across different regions.
```{r cv_training, echo=FALSE}
tr_data = training(model_input_split)
cv_tr_data <- spatial_block_cv(tr_data,
  v = 5,     
  cellsize = grid_cellsize(model_input),
  offset = grid_offset(model_input) + 0.00001
)
autoplot(cv_tr_data)
```

#### Building a Recipe
A recipe is the blueprint for the data handling and modeling process.

```{r recipe, echo=FALSE}
one_row_of_training_data = dplyr::slice(tr_data,1)
rec = recipe(one_row_of_training_data, formula = class ~ .)
rec
```


#### Creating a Workflow
My workflow will contain my recipe and a list of desired models. It is also used for specifying hyperparameters.
```{r workflow, echo=FALSE}
wflow = workflow_set(
  
  preproc = list(default = rec), # not much happening in our preprocessor
  
  models = list(                 # but we have 4 models to add
    
      # very simple - nothing to tune
      glm = logistic_reg(
          mode = "classification") |>
        set_engine("glm"),
      
      # two knobs to tune
      rf = rand_forest(
          mtry = tune(),
          trees = tune(),
          mode = "classification") |>
        set_engine("ranger", 
                   importance = "impurity"),
      
      # so many things to tune!
      btree = boost_tree(
          mtry = tune(), 
          trees = tune(), 
          tree_depth = tune(), 
          learn_rate = tune(), 
          loss_reduction = tune(), 
          stop_iter = tune(),
          mode = "classification") |>
        set_engine("xgboost"),
    
      # just two again
      maxent = maxent(
          feature_classes = tune(),
          regularization_multiplier = tune(),
          mode = "classification") |>
        set_engine("maxnet")
  )
)
wflow
```

Next, I chose the metrics for evaluating model performance. I used the default metrics from the tidysdm package (Boyce Index, TSS, and AUC) and included accuracy to provide an additional measure of performance.

```{r metrics, include=FALSE}
metrics = sdm_metric_set(yardstick::accuracy)
metrics
```


#### Fitting the Models to the Recipes

Firstly, I will find hyperparameters to use for fitting by iterating over different values for each.

```{r fit, warning = FALSE, include=FALSE}
wflow <- wflow |>
  workflow_map("tune_grid",
    resamples = cv_tr_data, 
    grid = 3,
    metrics = metrics, 
    verbose = TRUE)
```
Here's a quick plot of the outputs, with each metric being showcased.
```{r plot_wflow}
autoplot(wflow)
```


The goal is to choose the best set of hyperparameters for each model.

```{r select_best, include=FALSE}
model_fits = workflowset_selectomatic(wflow, model_input_split,
                                  filename = "Doryteuthis pealeii-v1-model_fits",
                                  path = data_path("models"))
model_fits
```



#### Exploring the Model Fit Results
We can make simple collective representations of the fitted models.

##### A table of metrics
We can quickly pull together summary statistics.

```{r model_fit_metrics, echo=FALSE}
model_fit_metrics(model_fits)
```


##### Confusion matrices and accuracy

We can plot confusion matrices and add accuracy. Accuracy is the proportion of the data that are predicted correctly.  
```{r con_matrices, echo=FALSE}
model_fit_confmat(model_fits)
```

##### ROC/AUC

These are plots of Receiver Operator Curves (ROC) from which the Area Under the Curve (AUC) is computed. The ROC illustrates the trade-off between the model's ability to correctly detect positives versus false alarms, while the Area Under the Curve (AUC) provides a single score representing the overall likelihood that the model can successfully distinguish between the two classes. The higher the AUC, the better.
```{r model_fit_roc_auc, echo=FALSE}
model_fit_roc_auc(model_fits)
```


My random forest and boosted regression tree models both have scores above 0.8! That's pretty exciting.

##### Variable importance

Variable importance can tell us about the contribution each covariate variable makes toward the whole.  

```{r model_fit_vip, echo=FALSE}
model_fit_varimp_plot(model_fits)
```

##### Partial Dependence Plots
To illustrate the relationship between each feature and the model's response, I plotted the partial dependence for every variable.

###### Boosted Tree
```{r pd_plot_bt, echo=FALSE}
model_fit_pdp(model_fits, wid = "default_btree", title = "Boosted Tree")
```


###### Generalized Linear Model
```{r pd_plot_glm, echo=FALSE}
model_fit_pdp(model_fits, wid = "default_glm", title = "Linear Model")
```


###### Random Forest
```{r pd_plot_rf, echo=FALSE}
model_fit_pdp(model_fits, wid = "default_rf", title = "Random Forest")
```


###### Maximum Entropy
```{r pd_plot_max, echo=FALSE}
model_fit_pdp(model_fits, wid = "default_maxent", title = "Max Entropy")
```



### Conclusion
After loading the data, I set aside a testing group to use only for the final check. I evaluated the models using standard metrics like accuracy and AUC, and I used partial dependence plots to visualize the impact of specific variables. 