---
title: "Predictions for My Second Species"
output: github_document
---
```{r knit, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
source("setup.R")
```
###### Zoe Ahmiegbe

## Introduction
For my second species, the shortfin squid, I will provide predictions of its presence under different scenarios (RCP 4.5 and RCP 8.5) for the present, 2055 and 2075.

RCP 45: This is an intermediate scenario where effective action is taken to reduce greenhouse gases.

RCP 85: This is the "worst-case" scenario (often called "business as usual") where emissions continue to rise unchecked through the end of the century .

### Nowcast: Present Time Period + Current Scenario
This section provides a nowcast (predictions about the present) according to the current climate conditions.

I will load in my Brickman data first and then read in my model fits from my previous exploration.
```{r load_covar, include=FALSE}
cfg = read_configuration(scientificname = "Illex illecebrosus",
                         version = "v2", 
                         path = data_path("models"))
db = brickman_database()
#db = brickman_database()
present_conditions = read_brickman(db |> filter(scenario == "PRESENT", 
                                                interval == "mon"),
                       add = c("depth", "month")) |>
  select(all_of(cfg$keep_vars))
```

```{r mod_fits, include=FALSE}
model_fits = read_model_fit(filename = "Illex illecebrosus-v1-model_fits")
model_fits
```

```{r nowcast, include=FALSE}
nowcast = predict_stars(model_fits, present_conditions)
nowcast
```

Now we can plot what is often called a "habitat suitability index" (hsi) map. We can 

```{r plot_nowcast_maxent, warning = FALSE, echo=FALSE}
plot_prediction(nowcast['default_btree'])+ ggtitle("Year 2025 - Current")
```


This is a presence/absence map with a threshold of 0.5, meaning that a presence means the prediction was greater than 0.5. Using a higher threshold would probably provide more accurate results.

```{r plot_class_labels_current, warning = FALSE, echo=FALSE}
pa_nowcast = threshold_prediction(nowcast)
plot_prediction(pa_nowcast['default_btree'])
```

### Year 2055

#### Year 2055 + Scenario RCP45
Let's try forecasting under RCP45 conditions in 2055.  First we load those parameters, then run the prediction and plot.

```{r load_2055_RCP45, warning = FALSE, include=FALSE}
covars_rcp45_2055 = read_brickman(db |> filter(scenario == "RCP45", 
                                               year == 2055, 
                                               interval == "mon"),
                                  add = c("depth", "month")) |>
  select(all_of(cfg$keep_vars))
```

```{r forecast_2055_RCP45, include=FALSE}
forecast_2055_RCP45 = predict_stars(model_fits, covars_rcp45_2055)
forecast_2055_RCP45
```
Plotting the prediction for Year 2055 with scenario RCP45:

```{r plot_forecast_2055_RCP45, echo=FALSE}
plot_prediction(forecast_2055_RCP45['default_btree']) + ggtitle("Year 2055- RCP 4.5")
```


#### Year 2055 + Scenario RCP85
Let's try forecasting under RCP85 conditions in 2055.  First we load those parameters, then run the prediction and plot.

```{r load_2055_RCP85, warning = FALSE, include=FALSE}
covars_rcp85_2055 = read_brickman(db |> filter(scenario == "RCP85", 
                                               year == 2055, 
                                               interval == "mon"),
                                  add = c("depth", "month")) |>
  select(all_of(cfg$keep_vars))
```

```{r forecast_2055_RCP85, echo=FALSE}
forecast_2055_RCP85 = predict_stars(model_fits, covars_rcp85_2055)
forecast_2055_RCP85
```

Plotting the prediction for Year 2055 with scenario RCP85:
```{r plot_forecast_2055_RCP85, echo=FALSE}
plot_prediction(forecast_2055_RCP85['default_btree']) + ggtitle("Year 2055- RCP 8.5")
```



### Year 2075 + Scenario RCP45
Predictions for year 2075 under RCP45 are presented here:

```{r load_2075_RCP45, warning = FALSE, include=FALSE}
covars_rcp45_2075 = read_brickman(db |> filter(scenario == "RCP45", 
                                               year == 2075, 
                                               interval == "mon"),
                                  add = c("depth", "month")) |>
  select(all_of(cfg$keep_vars))
```

```{r forecast_2075_RCP45, include=FALSE}
forecast_2075_RCP45 = predict_stars(model_fits, covars_rcp45_2075)
forecast_2075_RCP45
```

```{r plot_forecast_2075_RCP45, warning = FALSE, echo=FALSE}
plot_prediction(forecast_2075_RCP45['default_btree']) + ggtitle("Year 2075- RCP 4.5")
```



### Year 2075 + Scenario RCP85
Predictions for year 2075 under RCP85 are presented here:


```{r load_2075_RCP85, warning = FALSE, include=FALSE}
covars_rcp85_2075 = read_brickman(db |> filter(scenario == "RCP85", 
                                               year == 2075, 
                                               interval == "mon"),
                                  add = c("depth", "month")) |>
  select(all_of(cfg$keep_vars))
```

```{r forecast_2075_RCP85, include=FALSE}
forecast_2075_RCP85 = predict_stars(model_fits, covars_rcp45_2075)
forecast_2075_RCP85
```

```{r plot_forecast_2075_RCP85, warning = FALSE, echo=FALSE}
plot_prediction(forecast_2075_RCP85['default_btree']) + ggtitle("Year 2075- RCP 8.5")
```



## Saving the Predictions

It's easy to save the predictions (and read then back with `read_prediction()`).

```{r save_pred}
# make sure the output directory exists
path = make_path("predictions")

write_prediction(nowcast,
                 scientificname = cfg$scientificname,
                 version = cfg$version,
                 year = "CURRENT",
                 scenario = "CURRENT")

# Year 2055
write_prediction(forecast_2055_RCP45,
                 scientificname = cfg$scientificname,
                 version = cfg$version,
                 year = "2055",
                 scenario = "RCP45")

write_prediction(forecast_2055_RCP85,
                 scientificname = cfg$scientificname,
                 version = cfg$version,
                 year = "2055",
                 scenario = "RCP85")

# Year 2075
write_prediction(forecast_2075_RCP45,
                 scientificname = cfg$scientificname,
                 version = cfg$version,
                 year = "2075",
                 scenario = "RCP45")

write_prediction(forecast_2075_RCP85,
                 scientificname = cfg$scientificname,
                 version = cfg$version,
                 year = "2075",
                 scenario = "RCP85")
```

